### 2017-07-02 opsem.tex
Latest comments by KC and Kia addressed.
Fig. 4 redone.
RG Rules added.

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we will use to formalize the intuitions
presented in the previous section. Natural numbers, (shared) variables
and arithmetic expressions constitute the syntactic class of
expressions ($e$).  Commands ($c$) include $\cskip$, assignment
statements, transaction (\C{txn}) lexical blocks, and their sequential
and parallel composition. We let $T_i$ for $i \in \mathbb{N}$ range
over transaction identifiers. When it is evident we are referring to a
transaction, we use the number $i$ instead of $T_i$ for identification
(\eg in $\C{txn}\langle i \rangle$). For notational convenience, we
let $t$ range over both expressions and commands.

We define a small-step operational semantics for this language in
terms of an abstract machine that generates an execution trace ($\E$).
The first component of the trace is a set ($\A$) of \emph{effects},
where an effect ($\eta$) is a tuple of (a). the identifier ($T_i$) of
the transaction that generated this effect, (b). an indentifier ($j$)
that uniquely identifies this effect, (c). the operation ($\op$)
documented by this effect, which can be either a read (\C{RD(X)}), a
write (\C{WR(X)}), or a transaction commit (\C{COMMIT}), and (d). the
value, if any, associated with the operation (\eg, value read or value
written). Functions $\txn$, $\id$, $\oper$, and $\rval$ project each
of the four aforementioned individual components of the tuple.
In every step of the evaluation, the machine reduces a \txnimp\ term
by executing a read, write or commit operation, generating an effect,
and extending the trace. Since effects include transaction
identifiers, the semantics distinguishes between terms ($t$) of
different transactions. For example, $\txnbox{t}_i$ denotes a term $t$
inside a transaction $T_i$.  Evaluation contexts are also
appropriately marked. For example, $\ectx_i$ denotes the evaluation
context for a term inside $T_i$. The other component of an execution
trace is a visibility relation ($\visZ$) that establishes a visibility
property between effects among different transactions. The intent and
mechanics of $\visZ$ is described in the sequel.

Fundamental to our development is the notion of a trace invariant
($\I$). $\I$ is a function from traces ($\E$) to first-order logical
formulas ($\Prop$) that define well-formedness
constraints over traces.  The machine takes a step only if the
resulting trace satisfies the constraints imposed by $\I$. This
behaviour is captured by the auxiliary reduction rule
\rulelabel{E-Aux} that factors out the trace extension aspect of the
evaluation by abstracting away the operation-specific behaviour as a
function that generates an appropriate effect. We let $\mathcal{F}$
denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
generate a new effect and extend the trace ($\E = (\A,\visZ)$)
\emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
(i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
execution that runs to completion, every small-step preserves the
well-formedness of a trace, thus ensuring the invariance of $\I$.
Note that the semantics makes no assumptions about $\I$ other than its
type. As such, it can be instantiated with any trace-parametric
proposition that expresses constraints over the given trace. For
instance, consider the $\psi_{RC}$ specification from
\S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
is the following term:\vspace*{-8pt}

\begin{smathpar}
\begin{array}{l}
  \forall \eta_1,\eta_2,.\; \txn(\eta_1) = \C{Wd1}
  \conj \txn(\eta_2) = \C{Wd2} \\
  \hspace*{0.6in}\conj \C{Wd1} \neq \C{Wd2} \conj \eta_1 \hboar
  \eta_2 \Rightarrow \C{Wd1} \hboar \eta_2 \\
\end{array}
\end{smathpar}

\noindent It is easy to interpret the above specification in the
context of a trace $\E$ that captures an execution of the program in
Fig.~\ref{fig:motiv-eg-1}. Such a trace-parametric formula can be
used to instantiate the trace invariant $\I$ in Fig.~\ref{fig:txnimp}.
The resultant operational semantics describes an abstract machine that
gets stuck if an operation of \C{Wd2} is executed in a state that
incorporates some, but not all the effects (including the \C{COMMIT})
of \C{Wd1}.
% While the machine takes a step only if the constraints are
% satisfied, it neither defines nor explicitly assumes an oracle to
% check satisfaction.

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
$\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
below) is not a transitive relation, and hence lets us capture
finer-grained isolation properties than $\hbZ$, which we leverage in
our development.  The session order relation captures the sequential
order of operations within a transaction. In particular, it relates
two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
\txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
monotonically increasing identifiers to effects, as defined by the
$\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux} ($\maxId(\A)$
returns the maximum number identifying an effect in $\A$). Evaluation
contexts ($\ectx_i$) for transaction-bound terms are defined so as to
enforce a deterministic sequential order of execution within a
transaction, leading to a deterministic total order among effect ids,
which defines the session order relation. Visibility ($\visZ$) on the
other hand relates effects across concurrent transactions.
Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
$\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
during its execution, thus effecting its return value
($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
the value ($\rval$) of the write effect with the highest id among its
visible effects (this is made possible by appropriately defining
$\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
value of a read depends on what write effects it can witness. An
operation can only witness the effects of already concluded
operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions.

A more notable source of non-determinism, however, is the
\rulelabel{E-Aux} rule, which allows the machine to expose an
arbitrary subset ($S$) of existing effects ($\A$) to the incoming
operation. In other words, the machine is not obligated to reveal the
effects of all previous operations to an incoming operation. This
relaxation allows the abstract machine to model the semantics of
weakly-consistent data stores. For instance, operations issued to an
eventually consistent ({\sc ec}) replicated store could be dispatched
to different replicas whose states may not be in any well-defined
relationship. By allowing operations to witness arbitrary subsets of
the global state, the semantics models the weak visibility properties
of such stores; we elaborate on the implication of this style of
definition in \S\ref{sec:store-consistency}.  Stronger visibility
properties can be expressed by imposing well-formedness constraints
over $\visZ$ via the trace invariant ($\I$). Since the abstract
machine is obligated to satisfy $\I$ at every step of the execution,
operations are guaranteed to experience the level of isolation
specified by $\I$.  Thus, in executions that run to completion, the
abstract machine models a store that provides the required levels of
isolation.  Notably, the machine achieves this without defining an
operational semantics for isolation levels, instead solely relying on
their declarative characterization as trace well-formedness
constraints to enforce isolation guarantees.
\S\ref{sec:ansi-isolation} specifies various ANSI SQL isolation levels
stated as trace well-formedness constraints.

% TRACE RELATIONS
% ---------------
\input{trace-relations}

% However, a machine that lets operations witness arbitrary subset of
% the global state offers no isolation whatsoever. For example, it may
% allow a read operation to witness writes of an uncommitted
% transaction, violating RC isolation. Fortunately, our ability to
% express an isolation semantics as constraints over happens-before
% order through $\visZ$ and $\soZ$ relations, and the property of the
% abstract machine to be parametric over the trace invariant ($\I$),
% lets us solve this problem.
% In particular, we continue to define isolation semantics as
% constraints over $\visZ$ and $\soZ$, but confine their domain of
% interpretation to the given trace so that they now become trace
% well-formedness constraints. Well-formedness constraints can be
% combined into a trace invariant ($\I$).

As described previously, \rulelabel{E-Aux} abstracts away the
operation-specific behaviour of a machine step as a function ($\F$)
that accepts a set ($S$) of effects chosen by the machine to make
visible to the operation, interprets the operation w.r.t. $S$, and
returns an appropriate effect that encodes its return value. Rules
\rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
such functions for read, write and commit operations, respectively.
The effect returned by the function in each case includes its
transaction id ($T_i$) along with an arbitrarily chosen effect id
($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
$\rval$ for a write is the value being written, and for commit it is
$\bot$. In case of a read, the value read depends on how the read
operation chooses to interpret the given set ($S$) of visible effects.
The interpretation may depend on application semantics. For example, a
monotonically increasing counter application may choose to let a write
with the largest value determine the value of a read. To accommodate
multiple interpretations, the semantics is made parametric over an
interpretation function ($\interp{\cdot}$) that accepts a set of
effects and a variable name, and returns the value associated with the
variable. A straightforward interpretation function that chooses the
last write (i.e., write with largest id) is shown below:

\begin{smathpar}
\begin{array}{lcl}
  \isMax(S,\eta) & \Leftrightarrow &  \forall (\eta'\in S).  
  \kind(\eta') = \kind(\eta) \\
  & & \hspace*{0.4in}\Rightarrow \eta' = \eta \disj \id(\eta') < \id(\eta)\\

\interp{S}(X) & = & \C{if}\;(\exists (\eta \in S). \kind(\eta) = \C{WR}(X) 
  \wedge \isMax(S,\eta)) \\
  & & \C{then}\;\rval(\eta)\;\C{else}\;0\\
\end{array}
\end{smathpar}

\noindent Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} define
congruence properties for top-level terms and transaction-bound terms,
respectively. The rules and evaluation contexts ($\ectx$ and
$\ectx_i$) are defined such that only certain kinds of terms are
allowed at the top-level and inside a transaction. In particular, a
\txnimp program at the top-level can either be a transaction, or a
parallel composition of transactions. A command inside a \C{txn} block
can either be an assignment, or a sequential composition of
assignments.

\input{ansi-isolation}

>>>>

% Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
% \txnimp, a core language that we will use to formalize the intuitions
% presented in the previous section. Natural numbers, (shared) variables
% and arithmetic expressions constitute the syntactic class of
% expressions ($e$).  Commands ($c$) include $\cskip$, assignment
% statements, transaction (\C{txn}) lexical blocks, and their sequential
% and parallel composition. We let $T_i$ for $i \in \mathbb{N}$ range
% over transaction identifiers. When it is evident we are referring to a
% transaction, we use the number $i$ instead of $T_i$ for identification
% (\eg in $\C{txn}\langle i \rangle$). Like variables, transaction
% identifiers are globally accessible. For notational convenience, we
% let $t$ range over both expressions and commands.

% We define a small-step operational semantics for this language in
% terms of an abstract machine that generates an execution trace ($\E$).
% The first component of the trace is a set ($\A$) of \emph{effects},
% where an effect ($\eta$) is a tuple of (a). the identifier ($T_i$) of
% the transaction that generated this effect, (b). the unique
% identifier ($j$) of this effect, (c). the operation ($\op$)
% documented by this effect, which can be either a read (\C{RD(X)}), a
% write (\C{WR(X)}), or a transaction commit (\C{COMMIT}), and (d). the
% value, if any, associated with the operation (\eg, value read or value
% written). Functions $\txn$, $\id$, $\oper$, and $\rval$ project each
% of the four aforementioned individual components of the tuple.  In
% every step of the evaluation, the machine reduces a \txnimp\ term by
% executing a read, write or commit operation, generating an effect, and
% extending the trace. Since effects include transaction identifiers,
% the semantics distinguishes between terms ($t$) of different
% transactions. For example, $\txnbox{t}_i$ denotes a term $t$ inside a
% transaction $T_i$.  Evaluation contexts are also appropriately marked.
% For example, $\ectx_i$ denotes the evaluation context for a term
% inside $T_i$. The other component of an execution trace is a
% visibility relation ($\visZ$) that establishes a visibility property
% between effects among different transactions. The intent and mechanics
% of $\visZ$ is described in the sequel.

% Fundamental to our development is the notion of a trace invariant
% ($\I$). $\I$ is a function from traces ($\E$) to first-order logical
% formulas ($\Prop$) that define well-formedness
% constraints over traces.  The machine takes a step only if the
% resulting trace satisfies the constraints imposed by $\I$. This
% behaviour is captured by the auxiliary reduction rule
% \rulelabel{E-Aux} that factors out the trace extension aspect of the
% evaluation by abstracting away the operation-specific behaviour as a
% function that generates an appropriate effect. We let $\mathcal{F}$
% denote this function.  \rulelabel{E-Aux} uses $\mathcal{F}$ to
% generate a new effect and extend the trace ($\E = (\A,\visZ)$)
% \emph{only if} the well-formedness constraints imposed by $\I$ on $\E$
% (i.e., $\I(\E')$) are satisfied. Otherwise, it gets stuck. In an
% execution that runs to completion, every small-step preserves the
% well-formedness of a trace, thus ensuring the invariance of $\I$.
% Note that the semantics makes no assumptions about $\I$ other than its
% type. As such, it can be instantiated with any trace-parametric
% proposition that expresses constraints over the given trace. For
% instance, consider the $\psi_{RC}$ specification from
% \S\ref{sec:motivation}, but with bounded $T_1$ and $T_2$ instantiated
% with \C{Wd1} and \C{Wd2}, respectively. The instantiated specification
% is the following term:\vspace*{-8pt}

% \begin{smathpar}
% \begin{array}{l}
%   \forall \eta_1,\eta_2.\; \txn(\eta_1) = \C{Wd1}
%   \conj \txn(\eta_2) = \C{Wd2} \\
%   \hspace*{0.6in}\conj \C{Wd1} \neq \C{Wd2} \conj \eta_1 \hboar
%   \eta_2 \Rightarrow \C{Wd1} \hboar \eta_2 \\
% \end{array}
% \end{smathpar}

% \noindent It is easy to interpret the above specification in the
% context of a trace $\E$ that captures an execution of the program in
% Fig.~\ref{fig:motiv-eg-1}. Such a trace-parametric formula can be
% used to instantiate the trace invariant $\I$ in Fig.~\ref{fig:txnimp}.
% The resultant operational semantics describes an abstract machine that
% gets stuck if an operation of \C{Wd2} is executed in a state that
% incorporates some, but not all the effects (including the \C{COMMIT})
% of \C{Wd1}.
% % While the machine takes a step only if the constraints are
% % satisfied, it neither defines nor explicitly assumes an oracle to
% % check satisfaction.

% As described in \S\ref{sec:motivation}, the semantics of various
% isolation levels can be captured as constraints over the
% happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
% in our model, composed of more fundamental \emph{session order}
% ($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
% $\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
% below) is not a transitive relation, and hence lets us capture
% finer-grained isolation properties than $\hbZ$, which we leverage in
% our development.  The session order relation captures the sequential
% order of operations within a transaction. In particular, it relates
% two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
% \txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
% monotonically increasing identifiers to effects, as defined by the
% $\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux} ($\maxId(\A)$
% returns the maximum number identifying an effect in $\A$). Evaluation
% contexts ($\ectx_i$) for transaction-bound terms are defined so as to
% enforce a deterministic sequential order of execution within a
% transaction, leading to a deterministic total order among effect ids,
% which defines the session order relation. Visibility ($\visZ$) on the
% other hand relates effects across concurrent transactions.
% Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
% $\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
% during its execution, thus effecting its return value
% ($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
% the value ($\rval$) of the write effect with the highest id among its
% visible effects (this is made possible by appropriately defining
% $\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
% value of a read depends on what write effects it can witness. An
% operation can only witness the effects of already concluded
% operations, which varies between executions due to the
% non-deterministic order of evaluating the parallel composition of
% transactions.

% A more notable source of non-determinism, however, is the
% \rulelabel{E-Aux} rule, which allows the machine to expose an
% arbitrary subset ($S$) of existing effects ($\A$) to the incoming
% operation. In other words, the machine is not obligated to reveal the
% effects of all previous operations to an incoming operation. This
% relaxation allows the abstract machine to model the semantics of
% weakly-consistent data stores. For instance, operations issued to an
% eventually consistent ({\sc ec}) replicated store could be dispatched
% to different replicas whose states may not be in any well-defined
% relationship. By allowing operations to witness arbitrary subsets of
% the global state, the semantics models the weak visibility properties
% of such stores; we elaborate on the implication of this style of
% definition in \S\ref{sec:store-consistency}.  Stronger visibility
% properties can be expressed by imposing well-formedness constraints
% over $\visZ$ via the trace invariant ($\I$). Since the abstract
% machine is obligated to satisfy $\I$ at every step of the execution,
% operations are guaranteed to experience the level of isolation
% specified by $\I$.  Thus, in executions that run to completion, the
% abstract machine models a store that provides the required levels of
% isolation.  Notably, the machine achieves this without defining an
% operational semantics for isolation levels, instead solely relying on
% their declarative characterization as trace well-formedness
% constraints to enforce isolation guarantees.
% \S\ref{sec:ansi-isolation} specifies various ANSI SQL isolation levels
% stated as trace well-formedness constraints.

% % TRACE RELATIONS
% % ---------------
% \input{trace-relations}

% % However, a machine that lets operations witness arbitrary subset of
% % the global state offers no isolation whatsoever. For example, it may
% % allow a read operation to witness writes of an uncommitted
% % transaction, violating RC isolation. Fortunately, our ability to
% % express an isolation semantics as constraints over happens-before
% % order through $\visZ$ and $\soZ$ relations, and the property of the
% % abstract machine to be parametric over the trace invariant ($\I$),
% % lets us solve this problem.
% % In particular, we continue to define isolation semantics as
% % constraints over $\visZ$ and $\soZ$, but confine their domain of
% % interpretation to the given trace so that they now become trace
% % well-formedness constraints. Well-formedness constraints can be
% % combined into a trace invariant ($\I$).

% As described previously, \rulelabel{E-Aux} abstracts away the
% operation-specific behaviour of a machine step as a function ($\F$)
% that accepts a set ($S$) of effects chosen by the machine to make
% visible to the operation, interprets the operation w.r.t. $S$, and
% returns an appropriate effect that encodes its return value. Rules
% \rulelabel{E-Var}, \rulelabel{E-Asgn} and \rulelabel{E-Commit} define
% such functions for read, write and commit operations, respectively.
% The effect returned by the function in each case includes its
% transaction id ($T_i$) along with an arbitrarily chosen effect id
% ($j$) that is later verified to be unique in \rulelabel{E-Aux}. The
% $\rval$ for a write is the value being written, and for commit it is
% $\bot$. In case of a read, the value read depends on how the read
% operation chooses to interpret the given set ($S$) of visible effects.
% The interpretation may depend on application semantics. For example, a
% monotonically increasing counter application may choose to let a write
% with the largest value determine the value of a read. To accommodate
% multiple interpretations, the semantics is made parametric over an
% interpretation function ($\interp{\cdot}$) that accepts a set of
% effects and a variable name, and returns the value associated with the
% variable. A straightforward interpretation function that chooses the
% last write (i.e., write with largest id) is shown below:

% \begin{smathpar}
% \begin{array}{lcl}
%   \isMax(S,\eta) & \Leftrightarrow &  \forall (\eta'\in S).  
%   \kind(\eta') = \kind(\eta) \\
%   & & \hspace*{0.4in}\Rightarrow \eta' = \eta \disj \id(\eta') < \id(\eta)\\

% \interp{S}(X) & = & \C{if}\;(\exists (\eta \in S). \kind(\eta) = \C{WR}(X) 
%   \wedge \isMax(S,\eta)) \\
%   & & \C{then}\;\rval(\eta)\;\C{else}\;0\\
% \end{array}
% \end{smathpar}

% \noindent Rules \rulelabel{E-Top-Ctx} and \rulelabel{E-Txn-Ctx} define
% congruence properties for top-level terms and transaction-bound terms,
% respectively. The rules and evaluation contexts ($\ectx$ and
% $\ectx_i$) are defined such that only certain kinds of terms are
% allowed at the top-level and inside a transaction. In particular, a
% \txnimp program at the top-level can either be a transaction, or a
% parallel composition of transactions. A command inside a \C{txn} block
% can either be an assignment, or a sequential composition of
% assignments.

% \input{ansi-isolation}
